Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: Cross-cutting issue. Try to help.

\subsection{A Recipe for Sustainable Software, Philip E. Bourne.}

@@

\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}

@@

\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}

\note{lead: Daniel S. Katz}

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in one of the
keynotes, the Developing and Maintaining Software panel, and the Policy panel.

In the opening keynote, Phil Bourne suggested that perhaps sustainability can be defined
as the effort that happens to make the essential things continue. This leads to
having to decide what it is that we want to sustain, whether what we want to sustain is
valuable, and finally, who would care if it went away, and how does one measure
how much they care.

@@

Attendees suggested that a key question that the definition of sustainability is
one issue on which the community needs to agree, and that ideally, an initial
definition would be determined during the workshop, or at least some progress
would be made towards this goal.  Another topic that came up is what the
goal of sustainability is.  Perhaps it's reproducible science, or
persistence, or quality, or something else.  Similarly, some attendees want
to understand how success in sustainability is measured.  How does a set of
developers know when they have actually achieved sustainable software?
This led to a comment that sustainability should be addressed throughout the
full software life cycle.

@@

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned
that this is a rather ambiguous concept, and that the lack of an accepted
definition gets in the way of integrating the concept into software engineering.
They suggest that sustainability is a non-functional requirement, and that the
quality of software architectures determines sustainability.  They then
propose that sustainability is a measure of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and scalability.
Finally, they develop an architecture evaluation framework based on scenarios
that helps to illuminate how to measure quality/sustainability at
the architectural level.

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
be used to measure software sustainability, and though these questions might falsely lead
to yes or no answers, the complete set would determine a range of values for
sustainability. These question are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future, even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users? Do users have a mechanism to ask questions and to learn about the element?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling
Sustainability papers in the panel.
As shown in Table~\ref{tab:defining-sustainability}, for each paper he
discussed what was meant by software, if there was a definition of sustainability, and
what the approach was to either understand or evaluate sustainability.

\begin{table}[h]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al. \cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al. \cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al. \cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz and Proctor \cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt, et al. \cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al. \cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

>>>>

\subsection{A Recipe for Sustainable Software, Philip E. Bourne.} \label{sec:keynote1}

@@

\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith} \label{sec:keynote2}

@@

\section{Cross-cutting Issue: Defining Sustainability \scriptsize{\note{lead: Daniel S. Katz}}}  \label{sec:cross-cutting}

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in the first
keynote (Section~\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.

In the opening keynote, Philip Bourne suggested that perhaps sustainability can be defined
as the effort that happens to make the essential things continue. This leads to
having to decide what it is that we want to sustain, whether what we want to sustain is
valuable, and finally, who would care if it went away, and how one measures
how much they care.

@@

Attendees suggested that a key question that the definition of sustainability is
one issue on which the community needs to agree, and that ideally, an initial
definition would be determined during the workshop, or at least some progress
would be made towards this goal.  Another topic that came up is what the
goal of sustainability is.  Perhaps it is reproducible science, or
persistence, or quality, or something else.  Similarly, some attendees want
to understand how success in sustainability is measured.  How does a group of
developers know when they have actually achieved sustainable software?
This led to a comment that sustainability should be addressed throughout the
full software life cycle.

@@

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned
that this is a rather ambiguous concept, and that the lack of an accepted
definition gets in the way of integrating the concept into software engineering.
They suggest that sustainability is a non-functional requirement, and that the
quality of software architectures determines sustainability.  They then
propose that sustainability is a measure of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and scalability.
Finally, they develop an architecture evaluation framework based on scenarios
that help to illuminate how to measure quality or sustainability at
the architectural level.

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
be used to measure software sustainability, and though these questions might falsely lead
to yes or no answers, the complete set would determine a range of values for
sustainability. These question are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future, 
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users? 
\item Do users have a mechanism to ask questions and to learn about the element?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling
Sustainability papers in the panel.
As shown in Table~\ref{tab:defining-sustainability}, for each paper he
discussed what software meant, whether there was a definition of sustainability, and
what the approach was to either understand or evaluate sustainability.

\begin{table}[t]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}.}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al. \cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al. \cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al. \cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz and Proctor \cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt, et al. \cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al. \cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}
